---
title: "Unit 3 Project"
author: Lance Hiew
output: html_document
date: "2026-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
```

Below is a template to get you started on the Unit 3 Project. Make sure you edit the metadata above to include your name. For questions that include code, the template to produce `R` chunks is below. **Any code you write that produces an output must be fully visible in the HTML document you hand in!**

```{r}
library(tidyverse)
library(ggplot2)
library(GGally)
library(MASS)
library(klaR)
str(iris)
head(iris)
```

```{r}
pairs(iris[1:4], main = "Edgar Anderson's Iris Data", font.main = 4, pch = 19)
```

```{r}
pairs(iris[1:4], main = "Edgar Anderson's Iris Data", pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```

```{r}
plot(iris$Petal.Length, iris$Petal.Width, pch = c(23, 24, 25)[unclass(iris$Species)], bg = c("red", "green3", "blue")[unclass(iris$Species)],
    main = "Edgar Anderson's Iris Data")
```

```{r}
library(ggplot2)

ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) + geom_point() + labs(title = "Edgar Anderson's Iris Data",
    x = " Petal Length", y = "Petal Width") + theme_bw()
```

```{r}
library(GGally)

ggpairs(iris, columns = 1:4, aes(color = Species, alpha = 0.5)) + labs(title = "Edgar Anderson's Iris Data") + theme_bw()
```

```{r}
library(MASS)

data <- iris %>%
    dplyr::select(Sepal.Length:Petal.Width)

group <- iris %>%
    pull(Species)

lda(data, group)
```

```{r}
fit <- lda(data, group)

pred.fit <- predict(fit, iris[1:4])
predictions <- pred.fit$class

iris.table <- table(iris$Species, pred.fit$class)
iris.table
```



# Part I: Morphological Data

## Question 1:

```{r}
iris %>%
  group_by(Species) %>%
  summarise(
    Sepal.Length = mean(Sepal.Length),
    Sepal.Width  = mean(Sepal.Width),
    Petal.Length = mean(Petal.Length),
    Petal.Width  = mean(Petal.Width)
  )
```

The trait means for Setosa and Versicolor are very close with one another, while Virginica is different when it comes to sepal length. The petal length shows that all flowers have different numbers, but the speal width have Versicolor abd Virginca to be close with one another in numbers.

## Question 2
```{r}
##1 Sepal Length vs Sepal Width
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(title = "Sepal Length vs Sepal Width",
       x = "Sepal Length",
       y = "Sepal Width",
       color = "Species") +
  theme_bw()

##2. Sepal.width vs Sepal Length
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +
  geom_point() +
  labs(title = "Sepal Width vs Sepal Length",
       x = "Sepal Width", y = "Sepal Length",
       color = "Species") +
  theme_bw()

###3 Sepal length vs petal length
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point() +
  labs(title = "Sepal Length vs Petal Length",
       x = "Sepal Length",
       y = "Petal Length",
       color = "Species") +
  theme_bw()

###4 Petal Length vs Sepal Length 
ggplot(iris, aes(x = Petal.Length, y = Sepal.Length, color = Species)) +
  geom_point() +
  labs(title = "Petal Length vs Sepal Length",
       x = "Petal Length", y = "Sepal Length",
       color = "Species") +
  theme_bw()

###5 Sepal length vs petal width 
ggplot(iris, aes(x = Sepal.Length, y = Petal.Width, color = Species)) +
  geom_point() +
  labs(title = "Sepal Length vs Petal Width",
       x = "Sepal Length",
       y = "Petal Width",
       color = "Species") +
  theme_bw()

###6 Petal.Width vs Sepal.length
ggplot(iris, aes(x = Petal.Width, y = Sepal.Length, color = Species)) +
  geom_point() +
  labs(title = "Petal Width vs Sepal Length",
       x = "Petal Width", y = "Sepal Length",
       color = "Species") +
  theme_bw()

###7 Sepal Width vs Petal Length 
ggplot(iris, aes(x = Sepal.Width, y = Petal.Length, color = Species)) +
  geom_point() +
  labs(title = "Sepal Width vs Petal Length",
       x = "Sepal Width", y = "Petal Length",
       color = "Species") +
  theme_bw()

###8 Petal Length vs Sepal Width 
ggplot(iris, aes(x = Petal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(title = "Petal Length vs Sepal Width",
       x = "Petal Length", y = "Sepal Width",
       color = "Species") +
  theme_bw()

###9 Sepal Width vs Petal Width 
ggplot(iris, aes(x = Sepal.Width, y = Petal.Width, color = Species)) +
  geom_point() +
  labs(title = "Sepal Width vs Petal Width",
       x = "Sepal Width", y = "Petal Width",
       color = "Species") +
  theme_bw()

###10 Petal Width vs Sepal Width 
ggplot(iris, aes(x = Petal.Width, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(title = "Petal Width vs Sepal Width",
       x = "Petal Width", y = "Sepal Width",
       color = "Species") +
  theme_bw()

#11 Petal length vs Petal Width 
ggplot(iris, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point() +
  labs(title = "Petal Length vs Petal Width",
       x = "Petal Length", y = "Petal Width",
       color = "Species") +
  theme_bw()

##12 Petal Width vs Petal Length 
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, color = Species)) +
  geom_point() +
  labs(title = "Petal Width vs Petal Length",
       x = "Petal Width", y = "Petal Length",
       color = "Species") +
  theme_bw()
```

## Question 3

```{r}
library(MASS)

data1 <- iris %>%
  dplyr::select(Petal.Length, Petal.Width)

group <- iris %>%
  pull(Species)
fit1 <- lda(data1, group)
pred1 <- predict(fit1, data1)$class
table(iris$Species, pred1)
plot(fit1, col = c("red","green3","blue")[unclass(group)])
```

```{r}
data2 <- iris %>%
  dplyr::select(Petal.Length, Sepal.Length)

fit2 <- lda(data2, group)

pred2 <- predict(fit2, data2)$class

table(iris$Species, pred2)

plot(fit2, col = c("red","green3","blue")[unclass(group)])
```

The plots that give the best predictions are the "petal length and sepal length plot" and "petal length and petal width plot." The "petal length and petal width plot" have the lowest error rate with clear separation, while sepal length and petal length have stronger variance.

## Question 4

```{r}
table(iris$Species, pred2)
```
I am confident in Edgar Anderson's predictions as there are only 3 misclassified species and his data is so popular that it even got its own Wikipedia page. The confusion table helped to evaluate the 2 measurements.


# Part II: Geographical and temporal data 
```{r}
library(dismo)

get.gbif <- function(genus, species, na.rm = TRUE, crop = TRUE) {
    # GBIF query
    gps <- gbif(genus, species, geo = FALSE)
    # drop NA entries; note that this option is visited only with na.rm=TRUE
    if (na.rm) {
        gps <- subset(gps, !is.na(lon) & !is.na(lat))
    }
    if (crop) {
        keep <- c(which(colnames(gps) == "species"), which(colnames(gps) == "lat"), which(colnames(gps) == "lon"), which(colnames(gps) ==
            "month"))
        gps <- gps[keep]
    }
    return(gps)
}
```
```{r}
setosa <- get.gbif("Iris", "setosa")
versicolor <- get.gbif("Iris", "versicolor")
virginica <- get.gbif("Iris", "virginica")

iris_gps <- bind_rows(setosa, versicolor, virginica)

write_csv(iris_gps, file = "iris_gps.csv", col_names = TRUE)
```
```{r}
iris_gps <- read_csv("iris_gps.csv")
```

```{r}
library(maps)

ggplot2::theme_set(theme_bw())

world.map <- map_data(map = "world")
us.map <- map_data(map = "usa")


ggplot(world.map, aes(long, lat, group = group)) + geom_polygon(fill = "white", colour = "grey50") + coord_quickmap() + geom_point(data = iris_gps,
    mapping = aes(x = lon, y = lat, group = species, color = species), size = 0.5)
```
```{r}
ggplot(data = iris_gps, aes(x = month, y = after_stat(density), group = species, fill = species)) + geom_histogram(position = "dodge",
    binwidth = 1) + facet_wrap(~species, nrow = 3) + xlim(1, 12) + theme_minimal()
```

## Question 1
```{r}
Diceros <- get.gbif("Diceros", "bicornis")

Q3_gps <- bind_rows(Diceros)

library(maps)

ggplot2::theme_set(theme_bw())

world.map <- map_data(map = "world")
us.map <- map_data(map = "usa")

ggplot(world.map, aes(long, lat, group = group)) + geom_polygon(fill = "white", colour = "grey50") + coord_quickmap() + geom_point(data = Q3_gps,
    mapping = aes(x = lon, y = lat, group = species, color = species), size = 0.5)

ggplot(us.map, aes(long, lat, group = group)) + geom_polygon(fill = "white", colour = "grey50") + coord_quickmap(xlim = c(-120,-55), ylim = c(25,65)) + geom_point(data = Q3_gps, mapping = aes(x = lon, y = lat, group = species, color = species), size = 0.5) + labs(title = "Diceros bicornis")
```

I can see that the species is mainly located in south of Africa. It also has a little bit of occurance in the United States.

## Question 2
```{r}
ggplot(us.map, aes(long, lat, group = group)) + geom_polygon(fill = "white", colour = "grey50") + coord_quickmap(xlim = c(-120,-55), ylim = c(25,65)) + geom_point(data = iris_gps, mapping = aes(x = lon, y = lat, group = species, color = species), size = 0.5) + labs(title = "Iris")
```

## Question 3
```{r}
ggplot(us.map, aes(long, lat, group = group)) + geom_polygon(fill = "white", colour = "grey50") + coord_quickmap(xlim = c(-95, -60), ylim = c(20,45)) + geom_point(data = iris_gps, mapping = aes(x = lon, y = lat, group = species, color = species), size = 0.5) + labs(title = "Iris")
```
## Question 4

Looking at the map, I can see that most of the species are present in the eastern part of the United States. This exercise told me that all three species favor the environment in the east as it they are adapted to possibly its surroundings and climate. Virginica is mainly covering all of east of the U.S, Versicolor is mainly found in northeast, and Setosa is mainly located in the very north in Canada, possible favoring colder climates. Species are most likely suited for different climates or environments as Virginica may be more suited for more humid environments, while Setosa may prefer colder environments.

## Question 5
```{r}
ggplot(data = iris_gps, aes(x = month, y = after_stat(density), group = species, fill = species)) + geom_histogram(position = "dodge", binwidth = 1) + facet_wrap(~species, nrow = 3) + xlim(1,12) + theme_minimal() + labs(title = "The Iris Species Histogram")
```

It is hard to find the exact temporal isolation with this method, as observations are biased, where the recorded months shows when the person was there to make the observation. The recorded reproductive isolation can be manipulated based off of favors in certain timing or locations.

# Part III: Chromosomal data 

```{r}
iris_chrom_dataset <- read_csv("iris_chromosomes.csv", skip = 1)

head(iris_chrom_dataset)
```

```{r}
iris_chrom <- iris_chrom_dataset %>%
    dplyr::select(resolved_name, sporophytic, parsed_n) %>%
    separate(resolved_name, into = c("genus", "species"), sep = " ", remove = TRUE, extra = "drop")

head(iris_chrom)
```
```{r}
setosa_chrom <- iris_chrom %>%
    filter(species == "setosa")

setosa_chrom
```
```{r}
virginica_chrom <- iris_chrom %>%
    filter(species == "virginica")

virginica_chrom
```
```{r}
versicolor_chrom <- iris_chrom %>%
    filter(species == "versicolor")

versicolor_chrom
```
## Question 1

```{r}
library(tidyverse)
clean_iris_chrom <- iris_chrom %>%
  mutate(sporophytic_clean = str_replace_all(iris_chrom$sporophytic, "2n =", "")) %>%
  filter(!str_detect(sporophytic, "^c\\.")) %>%
  filter(!str_detect(sporophytic, "\\(")) %>%
  separate_rows(sporophytic, sep = ",") %>%
  mutate(sporophytic = str_trim(sporophytic)) %>%
  mutate(sporophytic = as.numeric(sporophytic)) %>%
  drop_na(sporophytic)
```
```{r}
Mode <- function(x) {
  x <- x[!is.na(x)]
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
```
```{r}
clean_iris_modes <- clean_iris_chrom %>%
  group_by(genus, species) %>%
  summarise(
    sporophytic_mode = Mode(sporophytic),
    .groups = "drop"
  )

clean_iris_modes %>% 
  print(n = 20)
```

## Question 2
```{r}
ggplot(clean_iris_chrom, aes(x = sporophytic)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 1,
    fill = "white",
    color = "grey50"
  ) +
  labs(
    title = "Sporophytic Chromosome Count in Iris Species",
    x = "Sporophytic Chromosome Count",
    y = "Density"
  ) +
  theme_minimal()
```

## Question 3
```{r}
clean_iris_modes %>% 
  slice(197, 199, 167)
```
The mode sporophytic chromosome value for each species are Setosa has 38 chromosomes, Virginica has 70 chromosomes, and Verssicolor has 108 chromosomes. I notice that they all have chromosomal variation. 

# Part IV: Producing a scientific argument

## Question 1
The three potential species are seperate species. Looking at the morpholoigcal data, it seems to be all separate, as the LDA only had an error rate of around 0.33. Looking at the geographical data, setosa was no where near the regions of both Virginica and Versicolor. This can possibly mean a divergence of species. With the chromosomal data, it is evident that Versicolor is a hybrid as it has 108 chromosomes which is the sum of Setosa with 38 chromosomes and Virginica with 70 chromosomes. It looks like the designations by Anderson were correct. The additional data of geographic, temporal, and chromosomal data add to the inital hypothesis, as I stated above explaning each one. It's a representation of three sepcies.

